Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-04-08 doc/manuscript_hkk_drought_sensitivity.Rmd
refining intro (and abstract)
@krishnaanujan , biggest change here is near complete re-write of paragraph on microenvironments. Please see what you think.

# Abstract (150 words) {-}
Droughts are increasingly impacting forests worldwide, threatening their critical carbon sink. 
Tree species adaptations to hot and dry conditions -- for example, dry season deciduousness in tropical seasonal forests -- might confer drought resistance, but such resistance may be variable across microenvironments and drought events. 
We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry forest in <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework--->to assess tree growth responses to two climatically distinct droughts. 
Species and individuals exhibited a wide range of drought responses within each drought and inconsistent responses across droughts. 
Deciduous species were more sensitive to wet-season drought than evergreen species, and this difference was amplified in wetter microclimates. 
Exposed trees were more sensitive, but greater water access through larger size or topographic location buffered growth declines. 
Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.

@@

In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@ipcc_climate_2023]. 
Understanding these impacts is an important research priority, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought, or hot/dry conditions anomalously departing from climatic means during any given time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @refs ], droughts with different characteristics may in fact have very different effects on tree growth [@refs].
<!--Here we define drought as hot/dry conditions anomalously departing from their long-term means during any given time of year. -->
<!-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  -->
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. 
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tree species in tropical dry forests respond to different types of droughts, including those occurring at different times relative to the regular seasonal cycle of wet and dry seasons. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Tree species have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with dry conditions, often leading to their classification as "drought tolerant" or "drought sensitive" [@refs].
In tropical seasonal forests in particular, trees have evolved to maximise growth and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021]. 
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation potentially and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky? Hulshof?].

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Individual growth responses to drought can depend on the specific microenvironments that trees experience, leading to varied atmospheric and hydrological stresses with distinct ecological consequences. 
Physiologically, trees experience drought stress as a combination of higher evaporative demand and higher thermal stress on leaves, and greater stress on roots because of low soil moisture. 
This results in greater biophysical stress on trees that have exposed crowns [@vinod_thermal_2023] or rely on more ephemeral (typically shallower) water sources that are severely diminished under drought [@refs].<!--I (KAT) made a substantive change here. It's not necessarily the case that shallow-rooted trees or trees in dry microenvironments suffer more, as they'd also tend to have adaptations to handle those conditions.--> 
Consequently, larger trees, with more exposed crowns, suffer more growth declines during drought [@bennett_larger_2015] (**see issue [#24](https://github.com/forestgeo/growth-precip-thailand/issues/24)**)<!--; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years-->. 
<!-- To some extent trees with belowground investment in larger and deeper root systems, or are in wetter locations have lower growth declines during drought because of continued access to water [].  -->
However, these responses can potentially be different across species or with drought severity.
<!--Larger trees have larger root systems, but do not necessarily access deeper water [@ref_from_Panama] -->
Trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
Indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
Finally, species strategies may constrain tree light and water microenvironments; e.g. evergreen species can covary with soil moisture [@kunert_leaf_2021], shade tolerant and understory species may be rarely exposed to hot/dry air. 
<!-- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.  -->
Whether drought characteristics interfere positively or negatively with these evolutionary and physiological strategies and whether it leads to consistent or heterogenous sensitivities across drought events is as yet unknown.  
<!--[from KAT: While all good content, I find this paragraph unnecessary/ distracting. If you'd really like to include content here, I'd work it into the methods.]
- In sum, these methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
- Methods need to capture species strategies, distributions across microenvironments, and be precise enough.  
- Manual dendrometer band timeseries can represent a wide range of species strategies with equipment precision necessary to detect signals. --> 
 <!-- Within species, size, exposure and location can affect sensitivity to drought.  
- There is theory and evidence that tree size itself matters.  
- Theoretically, we expect that greater height makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium, as described by Darcy's law [@mcdowell_darcy_2015; @fernández-de-uña_role_inpress].
- Height... [see refs in @vinod_thermal_2023; @olson_plant_2018; @couvreurWaterTransportTall2018].
- As tree size increases, leaves exert lower control over hydraulic resistance [@wolfe_leaves_2023].
- @chen_hydraulic_2022  
- There is theory and evidence that crown exposure matters [ @vinod_thermal_2023;  @refs_in_ vinod_thermal_2023; @scharnweber_confessions_2019].
- Microclimate buffering leads to cooler, moister understory air [@vinod_thermal_2023].
- Soils under closed canopies would also be cooler during hot times of the year [@lembrechts_global_2022].
- Reduced evaporative demand would also make them moister, and this might be added to by hydraulic redistribution.
- Trees with exposed crowns suffered significant crown dieback at greater rates in the 2012-16 CA drought [@ma_tree_2023].  
- rather than necessarily helping during drought, water access will shape the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate--->

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tropical tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and individual microenvironments. 
We test this by examining individual tree growth during two ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. 
Specifically, we test predictions that
i) both species and individuals respond differently to different droughts. 
ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
iii) Exposed trees are more sensitive to drought, but water availability, either through larger size or topographic location, can buffer growth declines.

>>>>

# Abstract (150 words) {-}
Droughts are increasingly impacting forests worldwide, threatening their critical carbon sink. 
Tree species adaptations to hot and dry conditions -- for example, dry season deciduousness in tropical seasonal forests -- might confer drought resistance, but such resistance may be variable across microenvironments and drought events. 
We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry forest in <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework--->to assess tree growth responses to two climatically distinct droughts. 
Species and individuals exhibited a wide range of drought responses within each drought and inconsistent responses across droughts. 
Deciduous species were more sensitive to early wet-season drought than evergreen species, and this difference was amplified in wetter microclimates. 
Exposed trees were more sensitive, but presumed greater water access through larger size or wetter topographic positions buffered growth declines. 
Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.

@@

Globally, intense drought is becoming more frequent, affecting forest demography, diversity, and  carbon (C) cycling [@ipcc_climate_2023]. 
However, our ability to predict drought impacts on tree woody growth and resultant ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022]. 
For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @refs ], droughts with different characteristics may in fact have very different effects on tree growth [@refs].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. 
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@refs].
In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky? Hulshof?].

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Finally, the microenvironments that individual trees experience influence growth responses to drought. <!--leading to varied atmospheric and hydrological stresses with distinct ecological consequences.-->
Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021] or acclimatized [@refs].
Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments.
Aboveground, tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems [@refs] and thereby potential to access more reliable water sources [@refs].
However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium<!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.

<!--Whether drought characteristics interfere positively or negatively with these evolutionary and physiological strategies and whether it leads to consistent or heterogenous sensitivities across drought events is as yet unknown.  -->

<!--HYPOTHESIS PARAGRAPH-->
We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats. 
We test this by examining individual tree growth during two ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. 
Specifically, we test predictions that
i) both species and individuals respond differently to different droughts. 
ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
iii) Exposed trees are more sensitive to drought, but water availability, either through larger size or topographic location, can buffer growth declines.
